---
title: 'data.table: World Population - Solutions'
author: "Saghir Bashir"
date: "`r format(Sys.time(), '%a %d %b %Y (%H:%M:%S)')`"
output:
  html_document:
    highlight: tango
    theme: united
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
      smooth_scroll: no
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## World Population Data

We will use a subset of the United Nations world population data[^UNdata]. 
The subset has 5 yearly population estimates from 1950 until 2015 by age groups for
both males and females. The variables in the dataset are as follows.

- `continent`: Continent
- `Location`: Country name
- `Time`: Year
- `AgeGrp`: Age group (e.g. 15-19, 20-14)
- `AgeGrpStart`: initial age of the age group
- `PopMale`: Total male population (thousands)
- `PopFemale`: Total female population (thousands)

## Exercise Questions

We want to answer the following questions with data:

1. How did the Portuguese population change between 1950 and 2015?
2. What was the total world population by sex in 1950 and 2015?
3. What was the total population by continent and sex in 1950 and 2015?
4. What was the total population by country and sex in 1950 and 2015?
5. How did the total population of Portugal change between 1950 and 2015?
6. Which countries had the highest and lowest proportion of females in 1950 
and 2015?
7. Discuss how the proportion of Portuguese females changed between 1950 and 2015?


## Read in the data

First load the `data.table` package (hint: use the `library()` function). Then
use the `fread()` function to read in the data file `World-Population.csv` in an
object called `un`. Finally look at the `un` dataset using the `str()` function.

```{r freadData}
library(data.table)
un <- fread("World-Population.csv")
un
str(un)
```

### Pre-processing the Data

Start by preparing the data for analysis:

1. Rename `continent` to `Continent`, `Location` to `Country` and `Time` 
to `Year` (hint: use the `setnames()` function). 
2. Create variable `PopF` as 1000 times `PopFemale` (hint use `:=`). 
3. Create variable `PopM` as 1000 times `PopMale`. 
4. Create variable `PopT` as total of males and females.
5. Drop variables `PopFemale` and `PopMale` (hint use `c("var1", "var2"):=NULL`).
6. Order the data by continent, country, year and age group (hint: use the `setkey()` function)
7. Look at the data (by typing `un` and using the `str()` function).

```{r preprocess}
setnames(un, c("continent", "Location", "Time"), c("Continent", "Country", "Year"))
un[, PopF := 1000*PopFemale]
un[, PopM := 1000*PopMale]
un[, PopT := PopF + PopM]
un[, c("PopFemale", "PopMale"):=NULL]
setkey(un, Continent, Country, Year, AgeGrp)
un
str(un)
```


## Exercise Answers

### Portugal: 1950 v 2017

> How did the Portuguese population change between 1950 and 2015?

1. Filter to select Portugal and the years 1950 and 2015 (hint use `%in% c(...)`)

**What do you notice about the variable `AgeGrp`?**

```{r PT19502017}
pt <- un[Country=="Portugal" & Year %in% c(1950, 2015)]
pt
```

### World Population

> What is the total world population by year?

1. Create a new object called `wp` to assign the total world population by year. 
2. Create `TotF`, `TotM and `Total` as the total for females, males and both (hint:
use `un[  .(TotF = sum(PopF), ... ), by = ...]`).
3. Look at the `wp`.

```{r WorldTotal}
wp <- un[, .(TotF = sum(PopF), TotM = sum(PopM), Total = sum(PopT)), by = Year]
wp
```


### Population by Continent

> What is the total population by continent and sex in 1950 and 2015?

1. Create a new object called `ct` to assign total population by continent. 
2. Only keep data for 1950 and 2015.
3. Create `TotF`, `TotM and `Total` as the total for females, males and both.
4. Look at the `ct`.


```{r ContinentTotal}
ct <- un[Year %in% c(1950, 2015),
         .(TotF = sum(PopF), TotM = sum(PopM), Total = sum(PopT)), 
         by = .(Continent, Year)]
ct
```

What changes do you observe?

### Population by Country

> What is the total population by country and sex in 1950 and 2015?

1. Create a new object called `cy` to assign total population by country. 
2. Use chaining for the following steps.
3. Only keep data for 1950 and 2015.
4. Create `TotF`, `TotM and `Total` as the total for females, males and both. Keep 
the variable `Continent` in the `by` part.
5. Look at the `cy`.

```{r CountryTotal}
cy <- un[Year %in% c(1950, 2015),
         .(TotF = sum(PopF), TotM = sum(PopM), Total = sum(PopT)), 
         by = .(Continent, Country, Year)][
           order(Continent, Country, Year)]
cy
```


### Portugal: 1950 to 2015

> How did the total population of Portugal change between 1950 and 2015?

1. Select Portugal in the `cy` dataset.
2. Look at other countries to see what is happening.

```{r PTchange}
cy[Country == "Portugal"]
```


### High & Low Female Proportion

> Which countries had the highest and lowest proportion of females in 1950 
and 2015?

1. Start with dataset `cy` to create a new dataset called `pp`.
2. Calculate two new variables called `PropF` and `PropM` for the proportion of females and males, respectively.
3. Order `PropF` from the lowest to the highest and select the first 10 observations
for both 1950 and 2015.
4. Order `PropF` from the highest to lowest and select the first 10 observations
for both 1950 and 2015.



```{r proportion}
pp <- cy[, `:=`(PropF=TotF/Total, PropM=TotM/Total)][
  , .(Country, Year, PropF)]

pp[order(Year, PropF)][Year==1950][1:10]
pp[order(Year, PropF)][Year==2015][1:10]

pp[order(Year, -PropF)][Year==1950][1:10]
pp[order(Year,- PropF)][Year==2015][1:10]
```


### Portuguese Females Over Time

> Discuss how the proportion of Portuguese females changed between 1950 and 2015?

1. Select Portugal from the `pp` dataset.

```{r ptChanges}
pp[Country=="Portugal"]
```

[^UNdata]: Source: https://population.un.org/wpp/Download/Standard/CSV/
[Dataset: "WPP2017_PopulationByAgeSex_Medium.csv](https://population.un.org/wpp/DVD/Files/1_Indicators%20(Standard)/CSV_FILES/WPP2017_TotalPopulationBySex.csv)
