---
title: "Exercise Sheet: Base R Concepts"

# Use letters for affiliations
author:
  - name: Saghir Bashir
#     affiliation: a
# address:
#   - code: a
#     address: ilustat \url{www.ilustat.com}

# For footer text  TODO(fold into template, allow free form two-authors)
# I will use it to place the copyright notice (Creative Comments Share Alike)
lead_author_surname: "[CC BY SA](https://creativecommons.org/licenses/by-sa/4.0/) [ilustat](http://ilustat.com/) $\\bullet$ [info@ilustat.com](mailto:info@ilustat.com)"

# Place DOI URL or CRAN Package URL here
doi: "Learn more at http://ilustat.com/resources/"

# Abstract
abstract: |
   This exercise sheet is designed to help you understand the key concepts that 
   are essential to learning and using R. The key requirement is to have R
   installed. Follow the instructions below to learn more. Have fun!
  
# Optional: Acknowledgements
acknowledgements: |
  The R project (https://www.r-project.org) for a great product.
  
# Optional: One or more keywords
# keywords:
#   - R
#   - Base R
#   - Statistics
#   - Data Science

# Paper size for document, values of letterpaper and a4paper
papersize: a4

# Font size of the document, values of 9pt (default), 10pt, 11pt and 12pt
fontsize: 9pt

# Optional: Force one-column layout, default is two-column
#one_column: true

# Optional: Enables lineno mode, but only if one_column mode is also true
#lineno: true

# Optional: Enable one-sided layout, default is two-sided
# one_sided: true

# Optional: Enable section numbering, default is unnumbered
numbersections: true

# Optional: Specify the depth of section number, default is 5
secnumdepth: 3

# Optional: Skip inserting final break between acknowledgements, default is false
skip_final_break: true

# Optional: Bibliography 
# bibliography: pinp

# Optional: Enable a 'Draft' watermark on the document
watermark: FALSE

# Customize footer, eg by referencing the vignette
footer_contents: "Exercise Sheet: Base R Concepts"

# Produce a pinp document
output: 
  pinp::pinp:
    fig_caption: yes
    collapse: true
    keep_tex: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
library(tidyverse)
library(knitr)
library(kableExtra)
options(width=45)
```

## Be Brave and Experiment 

Making mistakes and experimenting is an essential part of learning R.
You can recover from most mistakes (e.g. by restarting R). To do this 
"safely" start with a *fresh* R session without any other data loaded 
(otherwise you could lose it). 

## Start R

Double click in the R (or RStudio) icon to start it.

## Warning

If an R command is not complete then R will show a plus sign (`+`) prompt
on second and subsequent lines until the command syntax is correct.
```{r continuation, echo=TRUE, eval=FALSE}
+
```
To break out this, press the escape key (`ESC`).

## Exercises: The following are exercises for you try with R.

# Assign Values to an Object

## Assign numbers to objects:
```{r numObjects ec}
year <- 1995
cases <- 523
year
cases
```

## Assign characters to objects:
```{r charObjects ec}
country <- "Portugal"
country
```

## Warning: R is case sensitive

```{r caseSens1}
city <- "Coimbra"
CITY <- "Porto"
city
CITY
```


# R as a Calculator

## You can use R as a basic calculator
```{r calcBasic}
72 + (34*51 - 4982)/(18*45) 
sqrt(81)
9**2
```

## Trigonometric and logarithmic functions
```{r calcTrig}
# pi is a stored value
pi
sin(pi/3)
tan(pi/3)
log10(1000)
log(5.2)
exp(1)
```

## Rounding & combining
```{r calcFn}
signif(1512337, 2)
round(16.15643, 2)
round(exp(sqrt(log(514))), 2)
```

# Some More R Concepts

You can do some clever and useful things with using the assignment operator  `<-`: 
```{r assignBasic}
price <- 412.65
tax <- 94.91
total <- price + tax
total
```

## Vectors

The objects presented so far have all been scalars (single values). Working with
vectors is where R shines best as they are the basic building blocks of datasets. To create a vector we can use the `c()` (combine values into a vector) function.
```{r cVector}
vYear <- c(1995, 2000, 2005, 2010, 2015)
vCases <- c(523, 489, 511, 472, 439)
vCountry <- rep("Portugal", 5)
vYear
vCases
vCountry
```

## Vector operations

```{r operationsVecs}
# Calculate the total by adding the tax to net
net <- c(108.99, 291.42, 16.28, 62.29, 31.77)
tax <- c(22.89, 17.49, 0.98, 13.08, 6.67)
total <- net + tax
total
# Convert from degrees centrigrade to fahrenheit
tempC <- seq(-40, 40, 10)
tempF <- (9*tempC)/5 + 32
tempC
tempF
```

## Missing Values

Missing values are coded as `NA` in R. 
```{r MissingValues}
height <- c(NA, 1.73, 1.53, 1.67, 1.66, 1.81) 
weight <- c(63, 70, 95, 63, NA, 77)
bmi <- weight/(height**2)
height
weight
round(bmi, 1)
```

## Comments

It is useful to put human readable comments in your programs. These comments could 
help the future you when you go back to your program. R comments start with a hash 
sign (`#`). Everything after the hash to the end of the line will be ignored by R.
```{r comments}
# This comment line will be ignored when run.
weight # Weight in kg
height # Height in metres
# bmi - Body Mass Index
signif(bmi, 3)
```

## Managing Objects

Use function `ls()` to list the objects in your workspace. The `rm()` function removes
(deletes) them.
```{r lsrm}
ls()
rm(cases, city, CITY, country, year, tempF, tempC)
ls()
```

## R Functions 

What do the following functions do?
```{r functionUsage}
myNums <- c(454, 939, 740, 701, 394, 79, 30)
sort(myNums)
sort(myNums, decreasing = TRUE)
rank(myNums)
rev(myNums)
```

```{r sumstatFunctions}
# With missing values
naNums <- c(105, 81, 17, NA, 52, 394)
mean(naNums)
mean(naNums, na.rm = TRUE)
sd(naNums, na.rm = TRUE)
```

## The useful `summary()` function
```{r summaryFn}
summary(naNums)
```

## Pasting text and numbers

```{r funExamples}
Names <- c("Leo", "Iris")
Age <- c(4, 5)
Text <- paste0(Names, " is ", Age, " years old,")
Text
toupper(Text)
tolower(Text)
```

# Additional Materials

## Selecting elements from a vector

```{r selectElements}
colour <- c("red", "blue", "pink", "cyan", "gray")
colour[4]
colour[3:5]
colour[c(5, 1, 3)]
```


## Evaluating logical expressions

```{r logicalexp}
xscale <- c(7, 10, 9, 6, 1, 8, 2)
xscale > 8
xscale[xscale > 8 | xscale <=2]
```

# Summary

With these exercises you now have the building blocks to learn more.


